---
title: "PANORAMIC tutorial"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{PANORAMIC tutorial}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Introduction

Spatial omics studies often include multiple samples per condition.
PANORAMIC provides a pooled framework for cell-type colocalization
analysis across those samples, while accounting for both within-sample
uncertainty and between-sample heterogeneity.

This vignette walks through:

1.  Simulating a toy dataset with two "biological" groups.
2.  Running the PANORAMIC workflow (preparation, spatial statistics,
    pooling).
3.  Testing differential colocalization between groups.
4.  Interpreting tables and visualizations.

# Installation

```{r installation, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("panoramic")
```

# Load packages

```{r libraries, message=FALSE, warning=FALSE}
library(panoramic)
library(SpatialExperiment)
library(SummarizedExperiment)
library(S4Vectors)
library(BiocParallel)
library(dplyr)
library(ggplot2)
```

# Simulate example data

We simulate six samples with three cell types: `(A)`, `(B)`, `(C)`.
Group 1 has a mostly random spatial pattern. Group 2 is generated with
stronger central colocalization between the first two cell types.

```{r simulate_data, message=FALSE, warning=FALSE}
set.seed(123)
toy <- panoramic_simulate_dataset(
  n_group1 = 3,
  n_group2 = 3,
  n_cells_group1 = 200,
  n_cells_group2 = 350,
  group_labels = c("group1", "group2"),
  scenario_group1 = "random",
  scenario_group2 = "colocalized",
  seed = 123
)

spe_list <- toy$spe_list
design <- toy$design
design
```

Quick visual check of the simulated geometry:

```{r plot_samples, message=FALSE, warning=FALSE}
plot_df <- do.call(rbind, lapply(spe_list, function(spe) {
  coords <- SpatialExperiment::spatialCoords(spe)
  data.frame(
    x = coords[, 1],
    y = coords[, 2],
    cell_type = SummarizedExperiment::colData(spe)$cell_type,
    sample_id = SummarizedExperiment::colData(spe)$sample_id,
    stringsAsFactors = FALSE
  )
}))

ggplot(plot_df, aes(x = x, y = y, color = cell_type)) +
  geom_point(size = 0.7, alpha = 0.7) +
  facet_wrap(~ sample_id, nrow = 2) +
  coord_equal() +
  theme_bw() +
  labs(x = "x", y = "y", color = "Cell type")
```

# Run PANORAMIC workflow

PANORAMIC has three major analysis steps:

1.  `panoramic()`: prepares each sample and computes pairwise spatial
    summaries (here: cross-L statistics at selected radii).
2.  `panoramic_meta()`: performs random-effects pooling for each feature
    (cell-type pair + radius) within each group.
3.  `panoramic_compare_groups()`: tests whether pooled effects differ
    between groups.

Key parameters here:

-   `radii_um`: distances (microns) where colocalization is summarized.
-   `stat = "Lcross"`: cross-type Ripley's L statistic.
-   `nsim`: Loh bootstrap simulations per sample/pair.
-   `tau2 = "SJ"`: between-sample variance estimator in meta-analysis.
    Comes from the `metafor` package.

```{r run_workflow, message=FALSE, warning=FALSE}
se <- panoramic(
  spe_list = spe_list,
  design = design,
  cell_type = "cell_type",
  radii_um = 10,
  stat = "Lcross",
  nsim = 50,
  verbose = FALSE,
  BPPARAM = BiocParallel::SerialParam()
)

se_meta <- panoramic_meta(se, tau2 = "SJ", group_col = "group")
se_diff <- panoramic_compare_groups(se_meta, group1 = "group1", group2 = "group2")
```

# Interpret pooled and differential outputs

`rowData(se_diff)` contains one row per feature:

-   feature identity (`ct1`, `ct2`, `radius_um`)
-   pooled estimates and heterogeneity by group
-   differential comparison (`beta_diff`, `p_diff`, `fdr_diff`)

Biological interpretation:

-   positive `beta_diff` means stronger colocalization in `group2` vs
    `group1`
-   negative `beta_diff` means weaker colocalization in `group2`
-   small `fdr_diff` indicates a robust group difference after
    multiplicity control

```{r inspect_results, message=FALSE, warning=FALSE}
res <- as.data.frame(SummarizedExperiment::rowData(se_diff))
head(res[, c("ct1", "ct2", "radius_um", "beta_diff", "p_diff", "fdr_diff")])
res %>% dplyr::filter(fdr_diff < 0.05)
```

# Visualize results

Forest plot for one cell-type pair:

```{r forest_plot, message=FALSE, warning=FALSE}
plot_forest(
  se_meta,
  ct1 = "(A)",
  ct2 = "(B)",
  radius_um = 10,
  group_col = "group",
  show_heterogeneity = TRUE
)
```

Volcano plot overview:

```{r volcano_plot, message=FALSE, warning=FALSE}
plot_volcano(se_diff)
```

Network summary:

```{r network_plot, message=FALSE, warning=FALSE}
net <- create_spatial_network(
  se_diff,
  fdr_threshold = 0.2,
  directed = FALSE,
  leiden_resolution = 1.0
)
plot_spatial_network(net, layout = "fr", node_size_by = "degree")
```

# Related resources

For complementary spatial analysis workflows, see:

-   `spatstat` family (`spatstat.geom`, `spatstat.explore`) for
    point-process spatial statistics.
-   `SPIAT` for spatial histology and tumor microenvironment analyses.
-   `imcRtools` for imaging mass cytometry spatial workflows.
-   `spicyR` and `spatialFDA` for differential spatial localization
    testing in imaging data.

```{r session_info}
sessionInfo()
```
